New cardinality estimation algorithms for HyperLogLog sketches

### 2017-02-15 paper/paper.tex
fixed some more typos

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}. A first version of this paper was published there on April 17, 2016.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch. In particular, it is shown that it gives more precise cardinality estimates for union, intersection, or relative complements of two sets that are both represented by HyperLogLog sketches compared to the conventional technique using the inclusion-exclusion principle. All the new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

A more theoretically profound estimator for HyperLogLog sketches that does not depend on empirical data and significantly improves the estimation error is the historic inverse probability estimator \cite{Ting2014, Cohen2014}. It trades memory-efficiency for mergeability. The estimator needs to be continuously updated while inserting elements and the estimate depends on the element insertion order. Moreover, the estimator cannot be further used after merging two sketches, which limits its application to single data streams. If this restriction is acceptable, the self-learning bitmap \cite{Chen2011} 
could also be used, which provides a similar trade-off and also needs less space than the original HyperLogLog method to achieve the same precision.

@@

\subsection{Inequalities for the maximum likelihood estimate}
In the following we derive lower and upper bounds for $\symXEstimate$.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCountVariate_0\right)\cdot
\symHelper\!\left(
\symX\cdot
\frac{
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
{\symNumReg-\symCountVariate_0}\right)
-
\left(\symNumReg-\symCountVariate_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCountVariate_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
\log\!\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCountVariate_0}
{\symCountVariate_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}}.
\end{equation}
Using the monotonicity of $\symHelper$, the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
-
\left(\symNumReg-\symCountVariate_0\right)
\end{equation}
can be found, where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$ and
$\symRegValVariate_\text{max} := \max(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegValVariate'_\text{max}}
\log\!\left(
1+
\frac{\symNumReg-\symCountVariate_0}
{
2^{\symRegValVariate'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCountVariate_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}}.
\end{equation}
If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor and both are proportional to the harmonic mean of $2^{\symRegValVariate_1},\ldots,2^{\symRegValVariate_\symNumReg}$. Hence, consequent application of the maximum likelihood method would have directly suggested to use a cardinality estimator that is proportional to the harmonic mean without knowing the raw estimator \eqref{equ:raw_estimator} in advance.

@@

\subsection{Estimation error}
\label{sec:maximum_likelihood_estimation_error}
We have investigated the estimation error of the maximum likelihood estimation algorithm for the same HyperLogLog configurations as for the improved raw estimation algorithm in \cref{sec:corrected_raw_estimation_error}. 
\cref{fig:max_likelihood_estimation_error_12_20,fig:max_likelihood_estimation_error_8_24,fig:max_likelihood_estimation_error_16_16,fig:max_likelihood_estimation_error_22_10,fig:max_likelihood_estimation_error_12_52,fig:max_likelihood_estimation_error_12_14} show very similar results for same HyperLogLog parameters. 
What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities. The standard deviation of the relative error is also slightly better for the maximum likelihood estimator than for the improved raw estimator as shown in \cref{fig:stdev_comparison}. Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:max_likelihood_estimation_error_22_10}).

@@

The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both using \cref{alg:union}. Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$. Let  $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ be the value of the considered register in the HyperLogLog sketch representing $\symSetA$, $\symSetB$, and $\symSetX$, respectively. The corresponding values in sketches for $\symSetS_1$ and $\symSetS_2$ are given by
\begin{equation}
\symRegValVariate_1 = \max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
,\quad
\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right).
\end{equation}
Their joint cumulative probability function is given as
\begin{align}
\symProbability\!\left(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
\right)
&=
\symProbability\!\left(
\max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
\wedge
\max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\wedge
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\wedge
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right).
\end{align}
Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are pairwise disjoint. Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ can be described by \eqref{equ:register_value_distribution}. If we take into account that pairwise distinct elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
\begin{equation}
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
=\begin{cases}
0 & \symRegVal_1 < 0 \vee \symRegVal_2 < 0
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix}}{\symNumReg 2^{\symRegVal_1}}
-
\frac{\symPoissonRate_{\symSetBSuffix}}{\symNumReg 2^{\symRegVal_2}}
-
\frac{\symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\min(\symRegVal_1, \symRegVal_2)}}
}
& 0\leq\symRegVal_1 \leq \symRegRange \wedge 0\leq\symRegVal_2\leq\symRegRange
\\
e^{
-
\frac{\symPoissonRate_{\symSetBSuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_2}}
}
& 0\leq\symRegVal_2 \leq \symRegRange < \symRegVal_1
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_1}}
}
&  0\leq\symRegVal_1 \leq \symRegRange < \symRegVal_2
\\
1
&
\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2.
\end{cases}
\end{equation}

@@

\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
+
\textstyle\frac{1}{4}
\symDistanceMeasure^2
\textstyle\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
\right)
\end{equation}
where $\symDistanceMeasure = \frac{\symPoissonRate_\symSetASuffix
+\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
 is the Jaccard distance. Since $
\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
$ is always in the range $[0,\frac{1}{4}]$, the probability for equal register values can be bounded by
\begin{equation}
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
\right)
\lesssim
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\lesssim
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
+
\textstyle\frac{1}{16}
\symDistanceMeasure^2
\right).
\end{equation}
As shown in \cref{fig:equal_register_probability} the bounds are very close, especially for small Jaccard distances, where the probability can be well approximated by
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1-
\symAlpha_\infty\symDistanceMeasure.
\end{equation}
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{equal-register-probability}
\caption{The approximate probability range of equal register values as a function of the Jaccard distance.}
\label{fig:equal_register_probability}
\end{figure}
This dependency on the Jaccard distance is very similar to that of minwise hashing \eqref{equ:minwise_hahsing_probability}, which makes the
HyperLogLog algorithm an interesting candidate for locality-sensitive hashing with respect to the Jaccard similarity. The memory-efficiency of different hashing approaches can be measured by a storage factor that is the variance of the distance estimator multiplied by the number of bits used for storing the hash signature \cite{Li2011}. For a hash algorithm that maps two different sets to the same $\symNumBitsMinwiseHashing$-bit hash value with probability $1-\symA\symDistanceMeasure$ with some constant $\symA\in(0,1]$, the storage factor is given by $\symNumBitsMinwiseHashing\symDistanceMeasure\left(\frac{1}{\symA}-\symDistanceMeasure\right)$. 
Particularly, for the case $D=0.3$, this gives \num{6.72} for conventional minwise hashing with 32-bit hash values ($\symNumBitsMinwiseHashing=32$, $\symA\approx 1$), \num{0.51} for 1-bit minwise hashing ($\symNumBitsMinwiseHashing=1$, $\symA=\frac{1}{2}$), and  \num{1.96} for the HyperLogLog algorithm with 6-bit registers ($\symNumBitsMinwiseHashing=6$, $\symA=\symAlpha_\infty$). This means that the HyperLogLog algorithm is not as memory-efficient as 1-bit minwise hashing, but significantly better than 32-bit minwise hashing, when estimating the Jaccard distance by just counting equal registers. However, in contrast to 1-bit minwise hashing the HyperLogLog sketch contains additional information which allows estimating the cardinality or merging two HyperLogLog hash signatures. Furthermore, the method described in \cref{sec:cardinality_estimation_set_intersections} would allow a more accurate estimation of the Jaccard distance by using the estimates for intersection and union sizes. This could be used for additional more precise filtering when searching for similar items.

@@

>>>>

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}. A first version of this paper was published there on April 17, 2016.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of data sets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch. In particular, it is shown that it gives more precise cardinality estimates for union, intersection, or relative complements of two sets that are both represented by HyperLogLog sketches compared to the conventional technique using the inclusion-exclusion principle. All the new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

A more theoretically profound estimator for HyperLogLog sketches that does not depend on empirical data and significantly improves the estimation error is the historic inverse probability estimator \cite{Ting2014, Cohen2014}. It trades memory efficiency for mergeability. The estimator needs to be continuously updated while inserting elements and the estimate depends on the element insertion order. Moreover, the estimator cannot be further used after merging two sketches, which limits its application to single data streams. If this restriction is acceptable, the self-learning bitmap \cite{Chen2011} 
could also be used, which provides a similar trade-off and also needs less space than the original HyperLogLog method to achieve the same precision.

@@

\subsection{Inequalities for the maximum likelihood estimate}
In the following we derive lower and upper bounds for $\symXEstimate$.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCountVariate_0\right)\cdot
\symHelper\!\left(
\symX\cdot
\frac{
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
{\symNumReg-\symCountVariate_0}\right)
-
\left(\symNumReg-\symCountVariate_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCountVariate_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
\log\!\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCountVariate_0}
{\symCountVariate_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^{\symRegRange+1}}}.
\end{equation}
Using the monotonicity of $\symHelper$, the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
-
\left(\symNumReg-\symCountVariate_0\right)
\end{equation}
can be found, where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$ and
$\symRegValVariate_\text{max} := \max(\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace)$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegValVariate'_\text{max}}
\log\!\left(
1+
\frac{\symNumReg-\symCountVariate_0}
{
2^{\symRegValVariate'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCountVariate_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}}.
\end{equation}
If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor and both are proportional to the harmonic mean of $2^{\symRegValVariate_1},\ldots,2^{\symRegValVariate_\symNumReg}$. Hence, consequent application of the maximum likelihood method would have directly suggested to use a cardinality estimator that is proportional to the harmonic mean without knowing the raw estimator \eqref{equ:raw_estimator} in advance.

@@

\subsection{Estimation error}
\label{sec:maximum_likelihood_estimation_error}
We have investigated the estimation error of the maximum likelihood estimation algorithm for the same HyperLogLog configurations as for the improved raw estimation algorithm in \cref{sec:corrected_raw_estimation_error}. 
\cref{fig:max_likelihood_estimation_error_12_20,fig:max_likelihood_estimation_error_8_24,fig:max_likelihood_estimation_error_16_16,fig:max_likelihood_estimation_error_22_10,fig:max_likelihood_estimation_error_12_52,fig:max_likelihood_estimation_error_12_14} show very similar results for same HyperLogLog parameters. 
What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities. The standard deviation of the relative error is also slightly better for the maximum likelihood estimator than for the improved raw estimator as shown in \cref{fig:stdev_comparison}. Furthermore, contrary to the improved raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:max_likelihood_estimation_error_22_10}).

@@

The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both using \cref{alg:union}. Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$. Let  $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ be the value of the considered register in the HyperLogLog sketch representing $\symSetA$, $\symSetB$, and $\symSetX$, respectively. The corresponding values in sketches for $\symSetS_1$ and $\symSetS_2$ are given by
\begin{equation}
\symRegValVariate_1 = \max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
,\quad
\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right).
\end{equation}
Their joint cumulative probability function is given as
\begin{align}
\symProbability\!\left(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
\right)
&=
\symProbability\!\left(
\max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
\wedge
\max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\wedge
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\wedge
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right).
\end{align}
Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are pairwise disjoint. Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ can be described by \eqref{equ:register_value_distribution}. If we take into account that pairwise distinct elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
\begin{equation}
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
=\begin{cases}
0 & \symRegVal_1 < 0 \vee \symRegVal_2 < 0
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix}}{\symNumReg 2^{\symRegVal_1}}
-
\frac{\symPoissonRate_{\symSetBSuffix}}{\symNumReg 2^{\symRegVal_2}}
-
\frac{\symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\min(\symRegVal_1, \symRegVal_2)}}
}
& 0\leq\symRegVal_1 \leq \symRegRange \wedge 0\leq\symRegVal_2\leq\symRegRange
\\
e^{
-
\frac{\symPoissonRate_{\symSetBSuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_2}}
}
& 0\leq\symRegVal_2 \leq \symRegRange < \symRegVal_1
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_1}}
}
&  0\leq\symRegVal_1 \leq \symRegRange < \symRegVal_2
\\
1
&
\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2.
\end{cases}
\end{equation}

@@

\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
+
\textstyle\frac{1}{4}
\symDistanceMeasure^2
\textstyle\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
\right)
\end{equation}
where $\symDistanceMeasure = \frac{\symPoissonRate_\symSetASuffix
+\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
 is the Jaccard distance. Since $
\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
$ is always in the range $[0,\frac{1}{4}]$, the probability for equal register values can be bounded by
\begin{equation}
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
\right)
\lesssim
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\lesssim
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
+
\textstyle\frac{1}{16}
\symDistanceMeasure^2
\right).
\end{equation}
As shown in \cref{fig:equal_register_probability} the bounds are very close, especially for small Jaccard distances, where the probability can be well approximated by
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1-
\symAlpha_\infty\symDistanceMeasure.
\end{equation}
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{equal-register-probability}
\caption{The approximate probability range of equal register values as a function of the Jaccard distance.}
\label{fig:equal_register_probability}
\end{figure}
This dependency on the Jaccard distance is very similar to that of minwise hashing \eqref{equ:minwise_hahsing_probability}, which makes the
HyperLogLog algorithm an interesting candidate for locality-sensitive hashing with respect to the Jaccard similarity. The memory efficiency of different hashing approaches can be measured by a storage factor that is the variance of the distance estimator multiplied by the number of bits used for storing the hash signature \cite{Li2011}. For a hash algorithm that maps two different sets to the same $\symNumBitsMinwiseHashing$-bit hash value with probability $1-\symA\symDistanceMeasure$ with some constant $\symA\in(0,1]$, the storage factor is given by $\symNumBitsMinwiseHashing\symDistanceMeasure\left(\frac{1}{\symA}-\symDistanceMeasure\right)$. 
Particularly, for the case $D=0.3$, this gives \num{6.72} for conventional minwise hashing with 32-bit hash values ($\symNumBitsMinwiseHashing=32$, $\symA\approx 1$), \num{0.51} for 1-bit minwise hashing ($\symNumBitsMinwiseHashing=1$, $\symA=\frac{1}{2}$), and  \num{1.96} for the HyperLogLog algorithm with 6-bit registers ($\symNumBitsMinwiseHashing=6$, $\symA=\symAlpha_\infty$). This means that the HyperLogLog algorithm is not as memory-efficient as 1-bit minwise hashing, but significantly better than 32-bit minwise hashing, when estimating the Jaccard distance by just counting equal registers. However, in contrast to 1-bit minwise hashing the HyperLogLog sketch contains additional information which allows estimating the cardinality or merging two HyperLogLog hash signatures. Furthermore, the method described in \cref{sec:cardinality_estimation_set_intersections} would allow a more accurate estimation of the Jaccard distance by using the estimates for intersection and union sizes. This could be used for additional more precise filtering when searching for similar items.

@@

\frac{\left|\symError_\symHelper\right|}{2}
